---
output: _output.yml
---
\begin{center}
\huge \textbf{Bayesian Inference of F1 Driver Ability}

\Large STAT 405: Bayesian Statistics - Sohbat Sandhu (79661179)

\large 18 April 2026
\end{center}

```{r d1, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(knitr)
library(tidyverse)

lap_df <- read_csv("data/processed/lap_data.csv")
tele_df <- read_csv("data/processed/telemetry_data.csv")
```

\section{Introduction}

Formula One (F1) performance is shaped by a complex interaction between driver decision-making, car performance, tire degradation, and race conditions. Modern telemetry makes these interactions measurable at high resolution: instead of analyzing only lap time, we can study micro-sector time, where a lap is segmented into 100 equal-distance regions. A central challenge is that observed micro-sector times are not direct measurements of driver skill. They confound skill with team/car advantages, tire compound and wear, temperature, and race interruptions.

\subsection{Reasearch Question}

Using microsectors telemetry and lap summary data from the \textbf{2023 Abu Dhabi Grand Prix} (see Sections 6.3), we aim to infer latent driver ability and rank drivers with uncertainy while controlling for confounders and separaing driver effects from (1) team/car effects and (2) systematic track-geometry differences across microsectors. A bayesian hierarchical model is appropriate for this task as each microsector time is a noisy measurement and is influenced by multiple hierarchical factors with uneven sample sizes (influenced by driver choices like choosing different tire compounds and using different strategies).

\subsection{Literature Review}

[Van Kesteren and Bergkamp (2023)](https://doi.org/10.1515/jqas-2022-0021) developed a Bayesian multilevel rank-ordered logit model using \textit{race finishing positions over multiple races and seasons} to infer driver skill and constructor advantage, showing that constructor effects dominate variation in outcomes. [Bell et al. (2016)](https://doi.org/10.1515/jqas-2015-0050) apply a cross-classified multilevel model to historical race results, decomposing performance into driver and team components. More recently, [Cappello and Hoegh (2025)](https://doi.org/10.48550/arXiv.2512.00640) introduce a Bayesian state-space model to capture latent tire degradation and its impact on lap times, demonstrating the importance of dynamic latent processes in explaining performance variation. While these studies establish strong foundations, they primarily rely on coarse outcome variables such as finishing positions or lap times, or focus on single-driver dynamics. In contrast, this project differs by modelling micro-sector level telemetry, allowing inference on driver ability at a much finer spatial resolution.

\section{Preliminary Analysis}

We start with a preliminary analysis of the lap time data for 26 clean laps across 20 drivers (2 per team).

\begin{longtable}{l l c c c c l c c c c}
\caption{\textit{Team-Driver Lap Time Stats for 26 Clean laps (in s)}} \\ 
\vspace{-20pt} \\
\hline
\textbf{Team} & \textbf{D1} & \textbf{Mean} & \textbf{SD} & \textbf{Min.} & \textbf{q25, q75} & \textbf{D2} & \textbf{Mean} & \textbf{SD} & \textbf{Min.} & \textbf{q25, q75}\\
\hline
\endfirsthead

\multicolumn{11}{c}%
{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
\hline
\textbf{Team} & \textbf{D1} & \textbf{Mean} & \textbf{SD} & \textbf{Min.} & \textbf{q25, q75} & \textbf{D2} & \textbf{Mean} & \textbf{SD} & \textbf{Min.} & \textbf{q25, q75}\\
\hline
\endhead

\hline
\multicolumn{11}{r}{{Continued on next page}} \\
\endfoot

\hline
\endlastfoot
RedBull & VER & 88.932 & 1.153 & 86.993 & 87.91-89.57 & PER & 89.163 & 1.279 & 87.758 & 87.984-89.7 \\
Mercedes & HAM & 89.656 & 1.05 & 88.372 & 88.81-89.9 & RUS & 89.526 & 1.019 & 88.187 & 88.69-90.2 \\
Ferrari & LEC & 89.331 & 0.937 & 88.221 & 88.55-90.0 & SAI & 90.522 & 0.766 & 89.635 & 89.85-91.09 \\
McLaren & NOR & 89.467 & 0.942 & 88.164 & 88.60-89.87 & SAI & 89.731 & 1.172 & 88.149 & 88.74-90.59 \\
AstonMartin & ALO & 89.742 & 1.081 & 88.491 & 88.62-90.6 & STR & 89.891 & 1.129 & 88.177 & 88.86-90.85 \\
Alpine & GAS & 90.212 & 0.738 & 89.307 & 89.52-90.7 & OCO & 90.656 & 0.553 & 90.113 & 90.26-90.9 \\
Williams & ALB & 90.083 & 1.442 & 87.845 & 88.75-90.7 & SAR & 90.328 & 1.338 & 88.64 & 88.95-91.4 \\
AlphaTauri & TSU & 90.241 & 0.816 & 89.256 & 89.47-90.9 & RIC & 89.947 & 1.213 & 88.571 & 88.77-90.7 \\
AlfaRomeo & BOT & 91.41 & 0.885 & 89.998 & 90.57-92.0 & ZHO & 90.524 & 1.030 & 88.951 & 89.7-91.2 \\
Haas & MAG & 90.968 & 0.763 & 89.934 & 90.44-91.5 & SAI & 90.689 & 0.909 & 89.456 & 90.09-91.4
\end{longtable}
\vspace{-15pt}
\textit{Table 1} observes lap-level performance by driver. We observe that while there are differences in mean lap times across drivers and teams, there is also considerable overlap in their distributions, highlighting the need for a more nuanced analysis to disentangle driver ability from other factors.

```{r 1.1, echo=FALSE, out.width="49%", fig.show='hold', fig.align='center'}
knitr::include_graphics(
  c("figs/compound_comparison_by_team_group_row1.png",
    "figs/micro_sector_performance_heatmap.png")
)
```
\vspace{-20pt}
\textit{Figure 1.1: Lap time distributions by tire compound and team (every 2 drivers are a single team); Figure1.2: Heatmap of micro-sector performance across drivers and sectors.}

\textit{Figure 1.1} shows that lap times are faster with harder compounds, despite comparitively softer compounds generally yielding faster times. This suggest other factors like track temperature, tire degradation and stint length may be influencing performance. We also see a increase trend in lap times across teams, denoting that teams have different performance thresholds. \textit{Figure 1.2} reveals patterns in micro-sector performance, with consistent time increase corresponding to turns indicating the importance of reflecting track geometry. For example, sectors 26-31, 49-55, and 80-88 correspond to turns 5, 6-7, and 12-14 respectively (see \textit{Figure 6.4.1}), and show variability across drivers (hard to distinguish).

\section{Methodology}

\subsection{Modelling}

We model micro-sector and lap time based on logarithm scale to ensure predictions remain positive on the original scale and to improve approximate normality. We build two models: a lap level baseline (naive) model and a full hierarchical model (micro-sector level) for driver ranking and performance decomposition. All variables are specified from \textit{Sections 3.1.1-3.1.3} and implementation available in Section 6.2.

\subsubsection{Naive Model (Lap-Level Inferencing)}

Let $y_i = \log(\text{LapTimeSeconds}_i)$ for each observation $i$, we model $y_i \mid \mu_i, \sigma \sim \mathcal{N}(\mu_i, \sigma^2)$ where the performance is based on:
\[
\mu_i = \alpha_0 
+ \alpha_{d[i]} 
+ \beta_{c[i]} 
+ \beta_{\text{tyre}}\,\text{TyreLife}_i 
+ \beta_{\text{temp}}\,\text{TrackTemp}_i 
+ \beta_{\text{status}}\,\text{TrackStatus}_i
\]

Driver ability is included via a non-centered parameterization: $\alpha_d = \sigma_{\text{driver}} z_d, \ z_d \sim \mathcal{N}(0,1)$. Tyre compound effects are centered using $\beta_c$. For other priors we use $\beta_{\text{tyre}}, \beta_{\text{temp}}, \beta_{\text{status}} \sim \mathcal{N}(0,0.5)$. This model provides a baseline driver ranking adjusted for tyre degradation (based on usage, weather and compound) and acceptable racing conditions, without separating team/car contribution.

\subsubsection{Main Hierarchical Model (Micro-sector Level Inferencing)}

Let $y_i = \log(\text{TimeSeconds}_i)$ for each micro-sector observation $i$. We use a robust likelihood, $y_i \mid \mu_i, \sigma, \nu \sim t_\nu(\mu_i, \sigma)$ where the performance is based on:
\[
\mu_i = \alpha_0 
+ \alpha_{d[i]} 
+ \gamma_{t[i]} 
+ \delta_{m[i]} 
+ \theta_{i}
+ \beta_X^\top X_i 
+ \beta_{\text{status}}\,\text{TrackStatus}_i.
\]

Hierarchical effects with $z_d, z_t, z_m \sim \mathcal{N}(0,1)$:
\[
\alpha_d = \sigma_{\text{driver}} z_d \ \text{(Driver Effects)},
\gamma_t = \sigma_{\text{team}} z_t \ \text{(Team Effects)},
\delta_m = \sigma_{\text{micro}} z_m \ \text{(Micro-sector Effects)}
\]

Tyre degradation is accounted with compound effects ($\beta_c$), and the following track and tyre usage conditions:
$$\theta_{i} = \beta_{c[i]} + \beta_{\text{tyre}}\,\text{TyreLife}_i + \beta_{\text{temp}}\,\text{TrackTemp}_i + \beta_{\text{inter}}\,(\text{TyreLife}_i \cdot \text{TrackTemp}_i)$$
With regression priors as $\beta_{\text{tyre}}, \beta_{\text{temp}}, \beta_{\text{inter}}, \beta_X \ \text{(Driver Inputs)}, \beta_{\text{status}} \sim \mathcal{N}(0,0.5)$ This hierarchical model decomposes performance into driver ability, team/car effects, and track geometry, while accounting for tyre–environment interactions and driver inputs, enabling a more granular and fair estimation of latent driver skill.

\subsubsection{Prior Choice Motivations}

All continuous covariates (e.g., TyreLife, TrackTemp, driver inputs) are standardized, allowing the use of moderately informative priors with consistent interpretation across parameters. The intercept is assigned a weakly informative prior, $\alpha_0 \sim \mathcal{N}(0,2)$, for baseline log-time variation without permitting extreme values. Scale parameters are given exponential priors, $\sigma, \ \sigma_{\text{driver}}, \ \sigma_{\text{team}}, \ \sigma_{\text{micro}} \sim \text{Exp}(1)$, to ensure positive values. Hierarchical effects using $z \sim \mathcal{N}(0,1), \alpha = \sigma z$, improves sampling stability in high-dimensional spaces. Due to standardization, regression coefficients for tyre effects, environmental variables, and driver inputs use $\beta \sim \mathcal{N}(0,0.5)$, to indicate moderate effect on log-time variable by `1SD`. Categorical tyre compound effects are centered, $\beta_c = \beta_c^{\text{raw}} - \overline{\beta^{\text{raw}}}, \ \beta_c^{\text{raw}} \sim \mathcal{N}(0,0.5)$ to ensure performance is relative to the global average across all compound.The `df` parameter for the Student-$t$ likelihood is assigned $\nu \sim \mathcal{U}(2,60)$ to allow robustness to outliers which can be due to minor errors, resulting in significant differents in mirosector times. Overall, these priors are numerous and robust to leveraging information for decent inferences.

\section{Results \& Interpretations}

Posterior inference is performed using (i) NUTS MCMC via Stan for both models. Because the main model is computationally expensive and telemetry is high-volume, we additionally fit VI for the main model only as a faster posterior approximation. We used approximately, 410 observations (lap-level) for the naive model and 4400 observations (micro-sector-level subset) for the main model.

For both MCMC fits, diagnostic summaries for key driver-effect parameters and scale/separation parameters indicate acceptable convergence (i) $\hat{R}$ values are close to 1.0 and satisfy the practical threshold ($\hat{R} < 1.1$), (ii) both `ess_bulk` and `ess_tail` are sufficiently large relative to the retained sample size for stable posterior summaries of driver ability and separation parameters. (See variable specific summaries in Tables 6.4.1-6.4.2). Additionally for team/car contribution for varibles, the trace plots show good mixing and stationarity, with no divergent transitions or other warnings, indicating reliable sampling from the posterior distribution.

```{r d4.1, echo=FALSE, out.width="49%", fig.show='hold', fig.align='center'}
knitr::include_graphics(
  c("figs/diagnostics/main_trace_separation.png",
    "figs/diagnostics/main_rankhist_separation.png")
)
```
\vspace{-15pt}
\textit{Figure 4.1: Trace plots and rank histograms for separation parameters $(\sigma_{driver}, \sigma_{team})$ in the main hierarchical model, showing good mixing and no divergences.}

For the main model, VI posterior means are close to MCMC in central tendency for driver effects. In practice, this means posterior means/ranks are broadly similar, while VI behaves as a computationally efficient approximation under constrained runtime. As expected, VI can be slightly tighter in uncertainty than MCMC, so MCMC remains the reference posterior in interpretation.

```{r d4.2, echo=FALSE, out.width="45%", out.height="200px", fig.show='hold', fig.align='center'}
knitr::include_graphics(
  c("figs/diagnostics/main_driver_ability_mcmc_vs_vi.png",
    "figs/diagnostics/main_team_effect_mcmc_vs_vi.png")
)
```
\vspace{-15pt}
\textit{Figure 4.2: Variational inference vs MCMC for driver ability and team effect parameters in the main hierarchical model.}

\section{Limitations \& Conclusions}

The posterior distribution for the driver-team contribution fraction (driver variance (20%) share relative to driver & team variance) shows bimodal mass near the boundaries (peaks near low and high ends, see Figure 6.4.2). This suggests ambiguity in variance attribution between driver and team under the current model/data resolution. A notable substantive outcome is that the main model did not place VERSTAPPEN (Driver ID 1) as the top posterior driver, given the 2023 context, this indicates that the current hierarchical decomposition may be over-absorbing signal into team/track/interaction components or is sensitive to the reduced data slice used for feasible computation. Our telemetry data is huge, so we modeled only a subset, which may limit stability and representativeness. Complex interactions can blur driver vs team effects. We also ignore time-series dynamics. Finally, results are from one race (Abu Dhabi 2023), so season-wide generalization is limited.

\newpage <!-- APPENDIX -->

\section{Appendix}

\subsection{GitHub}

Outputs and code available at \href{https://github.com/sohbatSandhu/f1_telemetry_bayesian_analysis.git}{\textcolor{blue}{GitHub Repository}}

**NOTE:** To ensure no issues in running the code for reproducing the outputs and analysis follow the instructions to setup your environment and run instructions in the `README.md` file.

\subsection{Stan Implementation}

\subsubsection{Naive Model (lap-level inferencing)}

```{stan model1, message=FALSE, warning=FALSE, eval=FALSE, output.var="base_model"}
data {
  int<lower=1> N;
  int<lower=1> D;
  int<lower=1> C;

  array[N] int<lower=1,upper=D> driver;
  array[N] int<lower=1,upper=C> compound;
  array[N] int<lower=0,upper=1> TrackStatus;

  vector[N] y; // log_lap_time

  vector[N] TyreLife_z;
  vector[N] TrackTemp_z;
}

parameters {
  real alpha0;

  real<lower=1e-6> sigma_driver;
  vector[D] z_driver;

  vector[C] beta_comp_raw;

  real beta_tyre;
  real beta_temp;

  real beta_status;

  real<lower=1e-6> sigma;
}

transformed parameters {
  vector[D] alpha_driver = sigma_driver * z_driver;
  vector[C] beta_comp = beta_comp_raw - mean(beta_comp_raw);
}

model {
  alpha0 ~ normal(0, 2);

  // Use exponential to avoid piling mass at ~0, and keep sigma > 0.
  sigma ~ exponential(1);
  sigma_driver ~ exponential(1);

  z_driver ~ std_normal();

  beta_comp_raw ~ normal(0, 0.5);

  beta_tyre  ~ normal(0, 0.5);
  beta_temp  ~ normal(0, 0.5);

  beta_status ~ normal(0, 0.5);

  for (n in 1:N) {
    real mu = alpha0
      + alpha_driver[driver[n]]
      + beta_comp[compound[n]]
      + beta_tyre * TyreLife_z[n]
      + beta_temp * TrackTemp_z[n]
      + beta_status * TrackStatus[n];

    y[n] ~ normal(mu, sigma);
  }
}

generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    real mu = alpha0
      + alpha_driver[driver[n]]
      + beta_comp[compound[n]]
      + beta_tyre * TyreLife_z[n]
      + beta_temp * TrackTemp_z[n]
      + beta_status * TrackStatus[n];

    log_lik[n] = normal_lpdf(y[n] | mu, sigma);
  }
}

```

\subsubsection{Hierarchical Model (micro-sector level inferencing)}

```{stan model2, message=FALSE, warning=FALSE, eval=FALSE, output.var="main_model"}
data {
  int<lower=1> N;
  int<lower=1> D;
  int<lower=1> T;
  int<lower=1> M;
  int<lower=1> C;

  array[N] int<lower=1,upper=D> driver;
  array[N] int<lower=1,upper=T> team;
  array[N] int<lower=1,upper=M> micro;
  array[N] int<lower=1,upper=C> compound;

  array[N] int<lower=0,upper=1> TrackStatus;

  vector[N] y;

  vector[N] TyreLife_z;
  vector[N] TrackTemp_z;
  vector[N] TyreLife_x_TrackTemp;

  int<lower=1> K;
  matrix[N, K] X;

  int<lower=1> ppc_n;
  array[ppc_n] int<lower=1,upper=N> ppc_idx;
}

parameters {
  real alpha0;

  real<lower=1e-6> sigma_driver;
  vector[D] z_driver;

  real<lower=1e-6> sigma_team;
  vector[T] z_team;

  real<lower=1e-6> sigma_micro;
  vector[M] z_micro;

  vector[C] beta_comp_raw;

  real beta_tyre;
  real beta_temp;
  real beta_inter;

  vector[K] beta_X;
  real beta_status;

  real<lower=1e-6> sigma;
  real<lower=2, upper=60> nu;
}

transformed parameters {
  vector[D] alpha_driver = sigma_driver * z_driver;
  vector[T] gamma_team   = sigma_team * z_team;
  vector[M] delta_micro  = sigma_micro * z_micro;

  vector[C] beta_comp = beta_comp_raw - mean(beta_comp_raw);
}

model {
  alpha0 ~ normal(0, 2);

  // scale priors (more stable than half-normal at exactly 0)
  sigma ~ exponential(1);
  sigma_driver ~ exponential(1);
  sigma_team ~ exponential(1);
  sigma_micro ~ exponential(1);

  z_driver ~ std_normal();
  z_team ~ std_normal();
  z_micro ~ std_normal();

  beta_comp_raw ~ normal(0, 0.5);

  beta_tyre  ~ normal(0, 0.5);
  beta_temp  ~ normal(0, 0.5);
  beta_inter ~ normal(0, 0.5);

  beta_X ~ normal(0, 0.5);
  beta_status ~ normal(0, 0.5);

  nu ~ uniform(2, 60);

  for (n in 1:N) {
    real mu = alpha0
      + alpha_driver[driver[n]]
      + gamma_team[team[n]]
      + delta_micro[micro[n]]
      + beta_comp[compound[n]]
      + beta_tyre * TyreLife_z[n]
      + beta_temp * TrackTemp_z[n]
      + beta_inter * TyreLife_x_TrackTemp[n]
      + dot_product(beta_X, X[n])
      + beta_status * TrackStatus[n];

    y[n] ~ student_t(nu, mu, sigma);
  }
}

generated quantities {
  vector[N] log_lik;
  vector[ppc_n] y_rep;

  real driver_frac = square(sigma_driver) / (square(sigma_driver) + square(sigma_team));
  real team_frac   = 1 - driver_frac;

  for (n in 1:N) {
    real mu = alpha0
      + alpha_driver[driver[n]]
      + gamma_team[team[n]]
      + delta_micro[micro[n]]
      + beta_comp[compound[n]]
      + beta_tyre * TyreLife_z[n]
      + beta_temp * TrackTemp_z[n]
      + beta_inter * TyreLife_x_TrackTemp[n]
      + dot_product(beta_X, X[n])
      + beta_status * TrackStatus[n];

    log_lik[n] = student_t_lpdf(y[n] | nu, mu, sigma);
  }

  for (j in 1:ppc_n) {
    int n = ppc_idx[j];
    real mu = alpha0
      + alpha_driver[driver[n]]
      + gamma_team[team[n]]
      + delta_micro[micro[n]]
      + beta_comp[compound[n]]
      + beta_tyre * TyreLife_z[n]
      + beta_temp * TrackTemp_z[n]
      + beta_inter * TyreLife_x_TrackTemp[n]
      + dot_product(beta_X, X[n])
      + beta_status * TrackStatus[n];

    y_rep[j] = student_t_rng(nu, mu, sigma);
  }
}

```

\subsection{Data Information}

::: {.solbox data-latex=""}
\begin{longtable}{p{3cm} p{3cm} p{7cm}}

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\textit{Continued on next page}} \\
\midrule
\endfoot

\bottomrule
\endlastfoot

\multicolumn{3}{c}{\textbf{Car Telemetry Data across 100 Micro-sectors}} \\
\hline
\addlinespace

\texttt{Driver} & Categorical & Three-letter driver abbreviation (e.g., VER, HAM) \\
\texttt{Team} & Categorical & The team or constructor name (e.g., Red Bulls Racing, Mercedes, Ferrari) \\
\texttt{LapNumber} & Integer (1--58) & Sequential lap index during the race (Race for this session 58 laps long) \\
\texttt{micro\_sector} & Integer (0--99) & Track divided into 100 equal-distance segments \\
\texttt{Speed} & Continuous (km/h) & Instantaneous vehicle speed at given micro-sector \\
\texttt{Throttle} & Continuous (0--100) & Throttle application percentage \\
\texttt{Brake} & Continuous (0--100) & Brake application percentage \\
\texttt{RPM} & Continuous & Engine revolutions per minute \\
\texttt{TyreLife} & Integer & Number of laps completed on current tire set \\
\texttt{AirTemp} & Continuous ($^\circ$C) & Ambient air temperature \\
\texttt{TrackTemp} & Continuous ($^\circ$C) & Track temperature \\
\texttt{LapTimeSeconds} & Continuous (s) & Total lap time \\
\texttt{TimeSeconds} & Continuous (s) & Time elapsed from the start of the current micro-sector to its end \\

\addlinespace
\hline
\multicolumn{3}{c}{\textbf{Lap-Level Driver Summary Data}} \\
\hline
\addlinespace

\texttt{Driver} & Categorical & Three-letter driver abbreviation (e.g., VER, HAM) \\
\texttt{Team} & Categorical & The team or constructor name (e.g., Red Bulls Racing, Mercedes, Ferrari) \\
\texttt{LapNumber} & Integer (1--58) & Sequential lap index during the race (Race is 58 long) \\
\texttt{LapTimeSeconds} & Continuous (s) & Total lap time \\
\texttt{Stint} & Integer & Stint index between pit stops \\
\texttt{Compound} & Categorical & Tire compound (SOFT, MEDIUM, HARD) \\
\texttt{TrackStatus} & Binary & Race continuation status (1=Green, others=flags) \\
\texttt{PitInTime} & Date-time or NA & Timestamp of pit entry \\
\texttt{PitOutTime} & Date-time or NA & Timestamp of pit exit \\

\hline
\caption{\textit{Table Summary of Variables in the Telemetry for 26 Clean Laps}} \\
\end{longtable}
:::

\subsubsection{Processed Datasets}

The following are the previews of the two main datasets used in the analysis. The first dataset contains micro-sector telemetry data for 100 segments of each lap, while the second dataset summarizes lap-level information for 26 clean laps.

::: {.solbox data-latex=""}
```{r d2.1, echo=FALSE, warning=FALSE, message=FALSE}
glimpse(lap_df)
```
***Table 6.3.1:** Car Telemetry Data across 100 Micro-sectors driven by 20 drivers in 2023 F1 Abu Dhabi Grand Prix*
```{r d2.2, echo=FALSE, warning=FALSE, message=FALSE}
glimpse(tele_df)
```
***Table 6.3.2:** Lap Summary Data of 20 drivers (2 per team) in 2023 F1 Abu Dhabi Grand Prix*
:::

\subsection{Figures}

```{r d3.0.1, echo=FALSE, out.width="70%", fig.show='hold', fig.align='center'}
knitr::include_graphics(
  c("figs/yas_marina_image.png")
)
```
*Figure 6.4.1: Yas Marina Circuit Layout with turns labeled*

```{r d3.0.2, echo=FALSE, out.width="70%", fig.show='hold', fig.align='center'}
knitr::include_graphics(
  c("figs/diagnostics/driver_frac_density.png")
)
```
*Figure 6.4.2: Posterior distribution of driver-team contribution fraction (driver variance share relative to driver & team variance) in the main hierarchical model, showing bimodal mass near the boundaries, rather than a unimodal distribution at centered around 0.2*


:::{.solbox data-latex=""}
```{r d3.1, echo=FALSE, warning=FALSE, message=FALSE}
# Include the figures generated in from 07_diagnostics.R
main_key_summary <- read_csv("outputs/diagnostics/main_key_summary.csv")

# round values for better presentation
main_key_summary <- main_key_summary |>
  mutate(across(where(is.numeric), round, digits = 3)) |>
  select(variable, mean, q5, q95, rhat, ess_bulk, ess_tail)

# display the summary table
main_key_summary
```
*Table 6.4.1: Summary of key diagnostics for the MCMC main hierarchical model, including R-hat values, effective sample sizes, and posterior means for key parameters*

```{r d3.2, echo=FALSE, warning=FALSE, message=FALSE}
# Include the figures generated in from 07_diagnostics.R
main_driver_ability_summary <- read_csv("outputs/diagnostics/main_driver_ability_summary.csv")

# round values for better presentation
main_driver_ability_summary <- main_driver_ability_summary |>
  mutate(across(where(is.numeric), round, digits = 3)) |>
  select(variable, mean, q5, q95, rhat, ess_bulk, ess_tail)

# display the summary table
main_driver_ability_summary
```
*Table 6.4.2: Summary of driver ability diagnostics for the MCMC main hierarchical model, including R-hat values, effective sample sizes, and posterior means for key parameters.*

```{r d3.3, echo=FALSE, warning=FALSE, message=FALSE}
# Include the figures generated in from 07_diagnostics.R
naive_key_summary <- read_csv("outputs/diagnostics/naive_key_summary.csv")

# round values for better presentation
naive_key_summary <- naive_key_summary |>
  mutate(across(where(is.numeric), round, digits = 3)) |>
  select(variable, mean, q5, q95, rhat, ess_bulk, ess_tail)

# display the summary table
naive_key_summary
```
*Table 6.4.3: Summary of key diagnostics for the MCMC naive model, including R-hat values, effective sample sizes, and posterior means for key parameters.*

```{r d3.4, echo=FALSE, warning=FALSE, message=FALSE}
# Include the figures generated in from 07_diagnostics.R
naive_driver_ability_summary <- read_csv("outputs/diagnostics/naive_driver_ability_summary.csv")

# round values for better presentation
naive_driver_ability_summary <- naive_driver_ability_summary |>
  mutate(across(where(is.numeric), round, digits = 3)) |>
  select(variable, mean, q5, q95, rhat, ess_bulk, ess_tail)

# display the summary table
naive_driver_ability_summary
```
*Table 6.4.4: Summary of driver ability diagnostics for the MCMC naive model, including R-hat values, effective sample sizes, and posterior means for key parameters*
:::

\section{References}

1. Source: Open F1 - The open source API for Formula 1 data \textcolor{blue}{\url{https://openf1.org/}}
2. Source: Collected Primary Datasets - \textcolor{blue}{\url{https://github.com/sohbatSandhu/f1_telemetry_bayesian_analysis/tree/main/data/main}}
3. Source: Backup Datasets (1) Kaggle: \textcolor{blue}{\url{https://www.kaggle.com/datasets/gixarde31/f1-2023-abu-dhabi-gp-telemetry-and-lap-data}} (2) GitHub repository: \textcolor{blue}{\url{https://github.com/sohbatSandhu/f1_telemetry_bayesian_analysis/tree/main/data/backup}}
4. van Kesteren EJ, Bergkamp T. Bayesian analysis of Formula One race results: disentangling driver skill and constructor advantage. J Quant Anal Sports. 2023 Jul 25;19(4):273-293. doi: 10.1515/jqas-2022-0021. PMID: 38020583; PMCID: PMC10660124.
5. Bell, Andrew & Smith, James & Sabel, Clive & Jones, Kelvyn. (2016). Formula for success: Multilevel modelling of Formula One Driver and Constructor performance, 1950-2014. Journal of Quantitative Analysis in Sports. 12. 99-112. 10.1515/jqas-2015-0050. 
6. Cappello, C., & Hoegh, A. (2025). A state-space approach to modeling tire degradation in Formula 1 racing. arXiv:2512.00640. https://doi.org/10.48550/arXiv.2512.00640